Bayesian Hierarchical Modeling: An Introduction and Reassessment

### 2023-07-12 manuscript.Rmd
Remove blue text

The `R` package `brms` also allows fitting Bayesian models using `Stan`. However, model specification and model fitting can be achieved using a much simplified `R`-script \textcolor{blue}{alone}. Model specification is based on a formula syntax that is similar to the notation in the popular `lme4` package for frequentist hierarchical modeling [@R-lme4]. The formula is an object that specifies the dependent variable as a function of the independent variables. The `~` (tilde) separates the dependent variable (`rt`) on the left side from the independent variables on the right side. The formula distinguishes between general effects and individual deviations. The first part of the equation on the predictor side of the formula shown below (i.e., `1 + side + dif1 + dif2 + dif3 + dif4`) represents the general effects. The `1` represents the intercept, in the case of the symbolic distance effect $\mu_{\gamma}$, `side`, and `dif1` to `dif4` represent the general effects of side and distance ($\mu_{\beta}$ and $\mu_{\delta}$). Within the round brackets individual deviations and grouping variables are specified. In this case the grouping variable (`ind`) indicates that each participant has an individual effect. In other cases effects might vary per item, block, or even the combination of participant and item. The double bar (`||`) indicates that the correlation between the parameters should not be modeled. When using a single ba (`|`) then correlations across all individual effects (intercept, side and distance) are modeled. This approach, however, is a bit more difficult to interpret and would also not correspond to the implemented `Stan` model.

@@

```{r sdratioplotall, echo=FALSE, fig.cap="The prior and posterior distribution of the side parameter estimated by the \\textit{brms} package. The dots represent the height of the distributions at zero. The ratio of these dots indicates the Savage-Dickey density ratio. The Bayes factor represents the evidence for the hypothesis that the effect equals zero. A Bayes factor above one indicates that the evidence is in favor of the null hypothesis.", fig.align = "left", fig.height = 6, fig.width = 4, out.width="60%"}
knitr::include_graphics("I - Images/Fig17_sdrplotside_v2.png")
```
With `rstan`, the SDD-ratio can be computed by fitting a density function to the posterior distribution obtained from the MCMC samples, and comparing the value of this density function at $\mu_{\beta} = 0$ to the value of the prior distribution at $\mu_{\beta} = 0$. This is illustrated in the code block below. First, we extract the posterior samples for the parameter $\mu_{\beta}$, the general effect of side. We apply the `logspline` function [@logspline] to the samples to obtain an estimate of the log-density of the posterior distribution. The density at the zero is calculated with the `dlogspline` function. Finally, we compare the estimated density of the posterior to the density of the prior. \textcolor{blue}{This process is explained in further detail by Van Ravenzwaaij and Etz} [-@van2021simulation].

>>>>

The `R` package `brms` also allows fitting Bayesian models using `Stan`. However, model specification and model fitting can be achieved using a much simplified `R`-script \Revision{alone}. Model specification is based on a formula syntax that is similar to the notation in the popular `lme4` package for frequentist hierarchical modeling [@R-lme4]. The formula is an object that specifies the dependent variable as a function of the independent variables. The `~` (tilde) separates the dependent variable (`rt`) on the left side from the independent variables on the right side. The formula distinguishes between general effects and individual deviations. The first part of the equation on the predictor side of the formula shown below (i.e., `1 + side + dif1 + dif2 + dif3 + dif4`) represents the general effects. The `1` represents the intercept, in the case of the symbolic distance effect $\mu_{\gamma}$, `side`, and `dif1` to `dif4` represent the general effects of side and distance ($\mu_{\beta}$ and $\mu_{\delta}$). Within the round brackets individual deviations and grouping variables are specified. In this case the grouping variable (`ind`) indicates that each participant has an individual effect. In other cases effects might vary per item, block, or even the combination of participant and item. The double bar (`||`) indicates that the correlation between the parameters should not be modeled. When using a single ba (`|`) then correlations across all individual effects (intercept, side and distance) are modeled. This approach, however, is a bit more difficult to interpret and would also not correspond to the implemented `Stan` model.

@@

```{r sdratioplotall, echo=FALSE, fig.cap="The prior and posterior distribution of the side parameter estimated by the \\textit{brms} package. The dots represent the height of the distributions at zero. The ratio of these dots indicates the Savage-Dickey density ratio. The Bayes factor represents the evidence for the hypothesis that the effect equals zero. A Bayes factor above one indicates that the evidence is in favor of the null hypothesis.", fig.align = "left", fig.height = 6, fig.width = 4, out.width="60%"}
knitr::include_graphics("I - Images/Fig17_sdrplotside_v2.png")
```
With `rstan`, the SDD-ratio can be computed by fitting a density function to the posterior distribution obtained from the MCMC samples, and comparing the value of this density function at $\mu_{\beta} = 0$ to the value of the prior distribution at $\mu_{\beta} = 0$. This is illustrated in the code block below. First, we extract the posterior samples for the parameter $\mu_{\beta}$, the general effect of side. We apply the `logspline` function [@logspline] to the samples to obtain an estimate of the log-density of the posterior distribution. The density at the zero is calculated with the `dlogspline` function. Finally, we compare the estimated density of the posterior to the density of the prior. \Revision{This process is explained in further detail by Van Ravenzwaaij and Etz} [-@van2021simulation].

### 2023-07-12 manuscript.tex
Remove blue text

The \texttt{R} package \texttt{brms} also allows fitting Bayesian models using \texttt{Stan}. However, model specification and model fitting can be achieved using a much simplified \texttt{R}-script \textcolor{blue}{alone}. Model specification is based on a formula syntax that is similar to the notation in the popular \texttt{lme4} package for frequentist hierarchical modeling (Bates, Mächler, Bolker, \& Walker, 2015). The formula is an object that specifies the dependent variable as a function of the independent variables. The \texttt{\textasciitilde{}} (tilde) separates the dependent variable (\texttt{rt}) on the left side from the independent variables on the right side. The formula distinguishes between general effects and individual deviations. The first part of the equation on the predictor side of the formula shown below (i.e., \texttt{1\ +\ side\ +\ dif1\ +\ dif2\ +\ dif3\ +\ dif4}) represents the general effects. The \texttt{1} represents the intercept, in the case of the symbolic distance effect \(\mu_{\gamma}\), \texttt{side}, and \texttt{dif1} to \texttt{dif4} represent the general effects of side and distance (\(\mu_{\beta}\) and \(\mu_{\delta}\)). Within the round brackets individual deviations and grouping variables are specified. In this case the grouping variable (\texttt{ind}) indicates that each participant has an individual effect. In other cases effects might vary per item, block, or even the combination of participant and item. The double bar (\texttt{\textbar{}\textbar{}}) indicates that the correlation between the parameters should not be modeled. When using a single ba (\texttt{\textbar{}}) then correlations across all individual effects (intercept, side and distance) are modeled. This approach, however, is a bit more difficult to interpret and would also not correspond to the implemented \texttt{Stan} model.

@@

With \texttt{rstan}, the SDD-ratio can be computed by fitting a density function to the posterior distribution obtained from the MCMC samples, and comparing the value of this density function at \(\mu_{\beta} = 0\) to the value of the prior distribution at \(\mu_{\beta} = 0\). This is illustrated in the code block below. First, we extract the posterior samples for the parameter \(\mu_{\beta}\), the general effect of side. We apply the \texttt{logspline} function (Kooperberg, 2019) to the samples to obtain an estimate of the log-density of the posterior distribution. The density at the zero is calculated with the \texttt{dlogspline} function. Finally, we compare the estimated density of the posterior to the density of the prior. \textcolor{blue}{This process is explained in further detail by Van Ravenzwaaij and Etz} (2021).

>>>>

The \texttt{R} package \texttt{brms} also allows fitting Bayesian models using \texttt{Stan}. However, model specification and model fitting can be achieved using a much simplified \texttt{R}-script \textcolor{black}{alone}. Model specification is based on a formula syntax that is similar to the notation in the popular \texttt{lme4} package for frequentist hierarchical modeling (Bates, Mächler, Bolker, \& Walker, 2015). The formula is an object that specifies the dependent variable as a function of the independent variables. The \texttt{\textasciitilde{}} (tilde) separates the dependent variable (\texttt{rt}) on the left side from the independent variables on the right side. The formula distinguishes between general effects and individual deviations. The first part of the equation on the predictor side of the formula shown below (i.e., \texttt{1\ +\ side\ +\ dif1\ +\ dif2\ +\ dif3\ +\ dif4}) represents the general effects. The \texttt{1} represents the intercept, in the case of the symbolic distance effect \(\mu_{\gamma}\), \texttt{side}, and \texttt{dif1} to \texttt{dif4} represent the general effects of side and distance (\(\mu_{\beta}\) and \(\mu_{\delta}\)). Within the round brackets individual deviations and grouping variables are specified. In this case the grouping variable (\texttt{ind}) indicates that each participant has an individual effect. In other cases effects might vary per item, block, or even the combination of participant and item. The double bar (\texttt{\textbar{}\textbar{}}) indicates that the correlation between the parameters should not be modeled. When using a single ba (\texttt{\textbar{}}) then correlations across all individual effects (intercept, side and distance) are modeled. This approach, however, is a bit more difficult to interpret and would also not correspond to the implemented \texttt{Stan} model.

@@

With \texttt{rstan}, the SDD-ratio can be computed by fitting a density function to the posterior distribution obtained from the MCMC samples, and comparing the value of this density function at \(\mu_{\beta} = 0\) to the value of the prior distribution at \(\mu_{\beta} = 0\). This is illustrated in the code block below. First, we extract the posterior samples for the parameter \(\mu_{\beta}\), the general effect of side. We apply the \texttt{logspline} function (Kooperberg, 2019) to the samples to obtain an estimate of the log-density of the posterior distribution. The density at the zero is calculated with the \texttt{dlogspline} function. Finally, we compare the estimated density of the posterior to the density of the prior. \textcolor{black}{This process is explained in further detail by Van Ravenzwaaij and Etz} (2021).
